### 2016-09-08 Chapters/ResponseTheory.tex
Finishing up response theory chapter

Section \ref{sec:exact-response} will give a brief introduction to exact
state response theory for isolated molecules. The concepts of
quasienergy,\autocite{Christiansen1998-pe} variational perturbation
theory\autocite{Helgaker1992-ph} and pole-and-residue
analysis of response functions\autocite{Olsen1985-nr} will be introduced.
The exposition closely follows the one by \citeauthor{Saue2002-ns}
in \noparcite[ref.][]{Saue2002-ns}.
I will explicitly derive the \acrshort{SCF} response function for a
quantum/classical polarizable Hamiltonian in Section
\ref{sec:csm-response}. The derivation will employ the
density matrix-based, \acrshort{AO} formalism introduced by
\citeauthor{Thorvaldsen2008-sg} in \noparcite[ref.][]{Thorvaldsen2008-sg}.
This formalism is suitable for the derivation of arbitrary-order
response functions and their residues, also when perturbation-dependent
basis sets are considered.\autocite{Friese2015-kb, Friese2015-bu}
\paper{V} presents the arbitrary-order derivation of the \acrshort{PCM}.

\section{Exact State Response Theory}\label{sec:exact-response}

Let us consider the experimentally relevant situation where a molecular
system is subject to an external source of electromagnetic radiation.
Here and in the following, "external" will be used to mean an
electromagnetic field which is "weak" when compared to the
electron--electron and electron--nuclei interactions in the molecule.

@@

%\begin{subequations}
%  \begin{align}
%  \left.\deriv{\aveQ}{\epsilon_a}\right|_{\vect{\epsilon}=0} &= \Braket{0|V_A|0}\delta_{\omega_a} \\
%  \left.\frac{\diff^2 \aveQ}{\diff\epsilon_a\diff\epsilon_b}\right|_{\vect{\epsilon}=0}
%         &=
%         \response{V_A}{V_B}{\omega_b}\delta_{\omega_a+\omega_b} \\
%  \left.\frac{\diff^3\aveQ}{\diff\epsilon_a\diff\epsilon_b\diff\epsilon_c}\right|_{\vect{\epsilon}=0}
%  &=
%  \response{V_A}{V_B, V_C}{\omega_b,
%  \omega_c}\delta_{\omega_a+\omega_b+\omega_c}.
%  \end{align}
%\end{subequations}

@@

The perturbation-strength total derivative at zero perturbation strength
of the time-averaged quasi energy is:
\begin{equation}
\begin{aligned}
 \left[\prod_{i=1}^n \deriv{}{\epsilon_{x_i}}\right]\aveQ
 &= Q^{[n]}_{x_1 \ldots x_n}
 =
 Q^{[n]}_{0; x_1 \ldots x_n}
 + \sum_{j=1}^{n}E^{[n-1]}_{x_j;x_1\ldots x_{j-1}x_{j+1}\ldots x_n} \\
 &=
 \left[\prod_{i=1}^n \deriv{}{\epsilon_{x_i}}\right]Q_0
 + \sum_{j=1}^n
 \left[\prod_{i\neq j}^{n-1} \deriv{}{\epsilon_{x_i}}\right]E_{x_j}
\end{aligned}
\end{equation}
Notice that this is just an alternative and more verbose notation for
Eq. \eqref{eq:nth-order-derivative}, which allows to express response
functions in terms of wave function perturbed parameters.
%for first and second-order properties one has:
%\begin{subequations}
% \begin{align}
%   Q^{[1]}_{a} &= \left[Q^{[1]}_{0; a} + E_{a}\right]\delta_{\omega_a} =
%   \braket{0| V_A | 0}\delta_{\omega_a} \\
%  Q^{[2]}_{ab} &= \left[Q^{[2]}_{0; ab} + E^{[1]}_{a;b} + E^{[1]}_{b;a}\right]\delta_{\omega_a+\omega_b}
%  = \response{V_A}{V_B}{\omega_b}\delta_{\omega_a+\omega_b}
% \end{align}
%\end{subequations}
Application of the chain rule to the total derivatives in fact yields:
\begin{subequations}
 \begin{align}
   Q^{[1]}_{a} &=
   \sum_{\sigma}\left[
   Q^{[1]}_{0; \sigma}\lambda^{[1]}_{\sigma;a} + E_{a}
   \right]\delta_{\omega_a} \\
   Q^{[2]}_{ab} &=
   \sum_{\sigma\tau}\left[
   \lambda^{[1]}_{\sigma;a}Q^{[2]}_{0;\sigma\tau}\lambda^{[1]}_{\tau;b}
   + Q^{[1]}_{0; \sigma}\lambda^{[2]}_{\sigma;a}
   + E^{[1]}_{a;\sigma}\lambda^{[1]}_{\sigma;b}
   + E^{[1]}_{b;\sigma}\lambda^{[1]}_{\sigma;a}
   \right]\delta_{\omega_a+\omega_b} \label{eq:linres}
 \end{align}
\end{subequations}
where the following tensors were introduced:
\begin{subequations}
 \begin{align}
 Q^{[n]}_{0; \sigma_1\cdots \sigma_n} &=
 \frac{\partial^n Q_0}{\partial\lambda_{\sigma_1}\cdots\partial\lambda_{\sigma_n}} \\
 E^{[n]}_{x;\sigma_1\cdots \sigma_n} &=
 \frac{\partial^n E_x}{\partial\lambda_{\sigma_1}\cdots\partial\lambda_{\sigma_n}} \\
\lambda^{[n]}_{\sigma;x_1\dots x_n} &=
\frac{\partial^n \lambda_\sigma}{\partial \epsilon_{x_1}\cdots\partial \epsilon_{x_n}}
 \end{align}
\end{subequations}

@@

@@

\acrshort{PCM}~\autocite{Cammi1994-qj, Cammi1996-wf, Cammi1996-vx,
Cammi1999-rb, Cammi2003-qy, Frediani2005-nc, Ferrighi2010-pm}
\acrshort{FQ} and \acrshort{FQ}/\acrshort{PCM}~\autocite{Lipparini2012-hx, Lipparini2012-tl, Lipparini2013-ud}
\acrshort{PE}~\autocite{Olsen2010-wa}
\acrshort{QM}/\acrshort{MM}/Continuum~\autocite{Steindal2011-ki, Caprasecca2012-ir}
Open-ended \acrshort{AO}-based \acrshort{SCF} response~\autocite{
Larsen2000-hj, Kjaergaard2008-hy, Thorvaldsen2008-sg,
Kristensen2008-hv, Ringholm2014-gx}
Open-ended \acrshort{SCF} residues~\autocite{Friese2015-kb}
\paper{V}\autocite{pcm-openrsp}
Solution strategy for response equations~\autocite{Saad2003-oa,
Saad2011-gm, Kauczor2011-rd, Malmqvist2013-vw}

>>>>

Section \ref{sec:exact-response} will give a brief introduction to
response theory for isolated molecules. The concepts of
quasienergy,\autocite{Christiansen1998-pe} variational perturbation
theory\autocite{Helgaker1992-ph} and pole-and-residue
analysis of response functions\autocite{Olsen1985-nr} will be introduced.
The exposition closely follows the one by \citeauthor{Saue2002-ns}
in \noparcite[ref.][]{Saue2002-ns}.
I will explicitly derive the \acrshort{SCF} response function for a
quantum/classical polarizable Hamiltonian in Section
\ref{sec:csm-response}. The derivation will employ the
density matrix-based, \acrshort{AO} formalism introduced by
\citeauthor{Thorvaldsen2008-sg} in \noparcite[ref.][]{Thorvaldsen2008-sg}.
This formalism is suitable for the derivation of arbitrary-order
response functions and their residues, also when perturbation-dependent
basis sets are considered.\autocite{Friese2015-kb, Friese2015-bu}
\paper{V} presents the arbitrary-order derivation of the \acrshort{PCM}.

\section{Response Theory in a Nutshell}\label{sec:exact-response}

Let us consider the experimentally relevant situation where a molecular
system is subject to an external source of electromagnetic radiation,
Figure \ref{fig:OR}.
Here and in the following, "external" will be used to mean an
electromagnetic field which is "weak" when compared to the
electron--electron and electron--nuclei interactions in the molecule.

@@

@@

The perturbation-strength total derivative at zero perturbation strength
of the time-averaged quasi energy is:
\begin{equation}
\begin{aligned}
 \left[\prod_{i=1}^n \deriv{}{\epsilon_{x_i}}\right]\aveQ
 &= Q^{[n]}_{x_1 \ldots x_n}
 =
 Q^{[n]}_{0; x_1 \ldots x_n}
 + \sum_{j=1}^{n}E^{[n-1]}_{x_j;x_1\ldots x_{j-1}x_{j+1}\ldots x_n} \\
 &=
 \left[\prod_{i=1}^n \deriv{}{\epsilon_{x_i}}\right]Q_0
 + \sum_{j=1}^n
 \left[\prod_{i\neq j}^{n-1} \deriv{}{\epsilon_{x_i}}\right]E_{x_j}
\end{aligned}
\end{equation}
Notice that this is just an alternative and more verbose notation for
Eq. \eqref{eq:nth-order-derivative}, which allows to express response
functions in terms of wave function perturbed parameters.
Application of the chain rule to the total derivatives in fact yields:
\begin{subequations}
 \begin{align}
   Q^{[1]}_{a} &=
   \sum_{\sigma}\left[
   Q^{[1]}_{0; \sigma}\lambda^{[1]}_{\sigma;a} + E_{a}
   \right]\delta_{\omega_a} \\
   Q^{[2]}_{ab} &=
   \sum_{\sigma\tau}\left[
   \lambda^{[1]}_{\sigma;a}Q^{[2]}_{0;\sigma\tau}\lambda^{[1]}_{\tau;b}
   + Q^{[1]}_{0; \sigma}\lambda^{[2]}_{\sigma;a}
   + E^{[1]}_{a;\sigma}\lambda^{[1]}_{\sigma;b}
   + E^{[1]}_{b;\sigma}\lambda^{[1]}_{\sigma;a}
   \right]\delta_{\omega_a+\omega_b} \label{eq:linres}
 \end{align}
\end{subequations}
where the following tensors were introduced:
\begin{subequations}
 \begin{align}
 Q^{[n]}_{0; \sigma_1\cdots \sigma_n} &=
 \frac{\partial^n Q_0}{\partial\lambda_{\sigma_1}\cdots\partial\lambda_{\sigma_n}} \\
 E^{[n]}_{x;\sigma_1\cdots \sigma_n} &=
 \frac{\partial^n E_x}{\partial\lambda_{\sigma_1}\cdots\partial\lambda_{\sigma_n}} \\
\lambda^{[n]}_{\sigma;x_1\dots x_n} &=
\frac{\partial^n \lambda_\sigma}{\partial \epsilon_{x_1}\cdots\partial \epsilon_{x_n}}
 \end{align}
\end{subequations}

@@

The derivation above holds for an exact state, a situation that is never
realized in common practice. The power of response theory, however, is
in its applicability to exact and approximate state theories alike.
As noted by \citeauthor{Norman2011-ad}, a response theory treatment of
molecular properties can be summarized into the following four
steps:\autocite{Norman2011-ad}
\begin{enumerate}
    \item Single out a quantity of interest that can be connected to
      molecular properties. In our case, the time-averaged
      quasienergy;
    \item Introduce a suitable parametrization of the time-dependent,
      phase-separated wave function;
    \item Devise the appropriate equation of motion, based on the
      time-dependent Schr\"{o}dinger equation, for the time-dependent
      wave function parameters;
    \item Identify response functions and response equations.
\end{enumerate}
Thus, one can formulate and compute molecular properties based on
the approximate state ans\"{a}tze of quantum chemistry.
An interesting formulation of \acrshort{SCF} response theory is the
open-ended \acrshort{AO} density matrix-based approach presented by
\citeauthor{Thorvaldsen2008-sg}.\autocite{Thorvaldsen2008-sg}
By formulating the perturbation-strenght derivative of the time-averaged
quasienergy Lagrangian in terms of the \acrshort*{AO} density matrix,
the authors were able to formulate arbitrary order response functions
in the \acrshort*{AO} basis.
The density matrix is the variational degree of freedom and response
equations can be obtained by differentiation of the \gls{TDSCF}
equation, the stationarity condition on the density matrix.
The approach lends itself to an efficient \acrshort*{AO}-based
implementation,\autocite{Larsen2000-hj, Kjaergaard2008-hy} can fully
leverage different response parameters elimination
schemes,\autocite{Thorvaldsen2008-sg, Kristensen2008-hv} and is amenable
to an open-ended, recursive implementation.\autocite{Ringholm2014-gx,
Friese2015-kb, Friese2015-bu}
We will base our derivation of \acrshort*{SCF} response theory for
quantum/classical polarizable Hamiltonians of this framework.

@@

The development of response theory for quantum/classical polarizable
Hamiltonians has moved almost at the same pace as the polarizable
models themselves.
An ample literature exists in the context of the
\acrshort{PCM},\autocite{Cammi1994-qj, Cammi1996-wf, Cammi1996-vx,
Cammi1999-rb, Cammi2003-qy, Frediani2005-nc, Ferrighi2010-pm}
polarizable \acrshort{MM} models,\autocite{Curutchet2009-bt, Olsen2010-wa,
Lipparini2012-hx, Lipparini2012-tl}
and general \acrshort*{QM}/\acrshort*{MM}/Continuum models.\autocite{Steindal2011-ki, Caprasecca2012-ir, Lipparini2013-ud}

Solvent relaxation is an important point when formulating response
properties. Solvation dynamics occurs on timescales and through
relaxation modes that are typical of the solvent molecular structure and of the
peculiar solute-solvent interactions.
This is clearly difficult to model with continuum models, since the
entire frequency spectrum of the permittivity $\diel(\omega)$ would be
required.\autocite{Ingrosso2003-ev, Caricato2005-lo,
Mennucci2005-vi, Caricato2006-ba, Corni2015-pe}
However, if one consider the initial steps of an electronic excitation
process, a Franck--Condon-like response of the solvent can be assumed.
Nuclear motions inside and among solvent molecules will not be able to
follow the fast rearrangement of the solute electronic density, the
corresponding part of the response will remain frozen in the state
immediately previous the change.
The polarization vector will then be (approximately) a sum of two
components:
\begin{equation}
 \vect{P} \simeq \vect{P}_\mathrm{fast} + \vect{P}_\mathrm{slow},
\end{equation}
this is equivalent to consider only the asymptotic limits of the solvent
dielectric spectrum, \ie the static $\diels$ and optical
dielectric constants $\dield$.

The variational setting introduced in Section \ref{sec:variational},
offers a series of advantages also in the formulation of response
theory. As already noted, a classical counterpart to the
Hellmann--Feynman theorem holds:
\begin{equation}\label{eq:classical-HF}
  \deriv{U_\mathrm{pol}}{F} = \pderiv{U_\mathrm{pol}}{F}
  + \pderiv{U_\mathrm{pol}}{\s}\pderiv{\s}{F}
  = \pderiv{U_\mathrm{pol}}{F},
\end{equation}
which forms the basis for our formal development in
\paper{V}.\autocite{pcm-openrsp}
There we show how the variational polarization functional can
successfully achieve a transparent derivation of arbitrary order
response functions, when the fixed-cavity approximation for the
\acrshort*{PCM} is assumed.
Here we report a summary of our findings in \paper{V} employing the
classical functional in supermatrix form in Eq. \eqref{eq:supermatrix-functional}.
In the notation of \citeauthor{Thorvaldsen2008-sg}, we introduce the
generalized free energy:
\begin{equation}\label{eq:generalized-KS-energy}
  \begin{aligned}
 \tilde{\mathcal{G}}
 &=
 \tilde{\mathcal{E}}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p} +
 {}^t\tilde{\p}\Tr(\tilde{\mathbfup{s}}\pertmat{D}) \\
 &\eqtr
 \left[\pertmat{h} + \pertmat{V}^{t} +
 \frac{1}{2}\pertmat{G}^{\gamma}(\pertmat{D}) -
 \frac{\mathrm{i}}{2}\pertmat{T}\right]\pertmat{D}
 + \tilde{E}_\mathrm{xc}[\tilde{\rho}(\pertmat{D})] + h_\mathrm{nuc}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p}
 + {}^t\tilde{\p}\tilde{\mathbfup{s}}\pertmat{D},
 \end{aligned}
\end{equation}
whose perturbation-strength derivatives at zero perturbation strength
will yield the desired response functions.

Due to the implicit time-dependence of $\pertmat{D}$ and
$\tilde{\sigma}$, higher-order
derivatives of the KS generalized energy will require application of the
chain rule. The ${mn, abc\ldots}$ superscript describes how and to which
extent the chain rule was applied for a given term, \ie the number of explicit
differentiations with respect to the variational densities, so that
\begin{equation}\label{eq:differential}
  \mathcal{G}^{mn, abc}
  =
  \frac{\partial^{m+n+3}\mathcal{G}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c}
  =
    \mathcal{E}^{m, abc} +
  \frac{\partial^{m+n+3}U_\mathrm{pol}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c} \text{.}
\end{equation}
In this notation, the index $m$ denotes the order of differentiation
with respect to the density matrix $\mat{D}$, while the index $n$ symbolizes
the order of differentiation with respect to the \acs{ASC} density $\sigma$.
Differentiation with respect to the density matrix will result in a
$2m$-rank tensor, while differentiation with respect to the
\acs{ASC} density will result in a function of the continuous cavity
index $\vect{s}$. For higher-order properties, mixed terms involving both density
matrix and \acs{ASC} density differentiation may generally occur.
In the \emph{fixed-cavity approximation}, the cavity is kept frozen at a
given molecular geometry.\autocite{Cammi1994-qj}
Under this simplifying assumption, only the linear interaction term in
the polarization functional eq.~\ref{eq:PCM-functional} will be affected
by the movements of the nuclei \emph{via} the dependence of basis functions on
the molecular geometry.
Its perturbation-strength derivative will then be
\begin{equation}\label{eq:PCM-derivative}
 \deriv{}{\epsilon_a}\left\lbrace
 U_\mathrm{pol}
 \right\rbrace_T
 \eqavetr
 {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D} \text{,}
\end{equation}
where the second term only involves derivatives of the electrostatic potential integrals.
We remark that, under the fixed-cavity approximation, both the density
matrix -- $m$ -- and \acs{ASC} density -- $n$ -- differentiation indices
in eq.~\ref{eq:differential} can only assume the values 0 or 1 for the
$\frac{\partial^{m+n+3}U_\mathrm{pol}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c}$
term.
By construction, the density matrix dependence in the
polarization functional is at most linear, while,
by virtue of the classical Hellmann--Feynman theorem,
eq.~\ref{eq:classical-HF}, the \acs{ASC} variational density will also appear
at most \emph{linearly} in $\tilde{\mathcal{G}}^{00,a}$.
The free energy term perturbation strength derivative is given as
\begin{equation}\label{eq:tilde-G00-a}
  \begin{aligned}
    \tilde{\mathcal{G}}^{00,a}(\tilde{\mat{D}}, \tilde{\sigma}, t)
    &= \tilde{\mathcal{E}}^{0,a} +
    {}^t\tilde{\p}\Tr({\tilde{\mathbfup{s}}^a\pertmat{D}}) \\
  &\eqtr
  (\pertmat{h}^a + \pertmat{V}^{t,a} + \frac{1}{2}\pertmat{G}^{\gamma, a}(\pertmat{D}) + \pertmat{F}^{\Omega_a}_\mathrm{xc}
  -\frac{\mathrm{i}}{2}\pertmat{T}^a)\pertmat{D}
  + h_\mathrm{nuc}^a + {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D}
  \end{aligned}
\end{equation}
where the matrix $\pertmat{F}^{\Omega_a}_\mathrm{xc}$ denotes the
the functional derivative matrix defined in terms of the perturbed
overlap distributions $\tilde{\Omega}^a$.

Response functions can then be obtained by straightforward
differentiation with respect to further perturbations and subsequent
evaluation at zero perturbation strength, so that
\begin{subequations}
  \begin{align}
    L^a &\eqavetr
    \mathcal{G}^{00,a} - \mat{S}^a\mat{W} \label{eq:first-order} \\
    L^{ab} &\eqavetr \mathcal{G}^{00,ab}
    + \mathcal{G}^{10,a}\mat{D}^{b}
    + \mathcal{G}^{01,a}\p^{b}
    - \mat{S}^{ab}\mat{W}
    - \mat{S}^{a}\mat{W}^{b} \label{eq:second-order} \\
    L^{abc} &\eqavetr
      \mathcal{G}^{00,abc}
    + \mathcal{G}^{10,ac}\mat{D}^{b}
    + \mathcal{G}^{10,ab}\mat{D}^{c}
    + \mathcal{G}^{20,a}\mat{D}^{b}\mat{D}^{c} \nonumber \\
    &+ \mathcal{G}^{10,a}\mat{D}^{bc}
    + \mathcal{G}^{11,a}\mat{D}^{b}\p^c \label{eq:third-order} \\
    &+ \mathcal{G}^{01,ac}\p^{b}
    + \mathcal{G}^{01,ab}\p^{c}
    + \mathcal{G}^{01,a}\p^{bc}
    + \mathcal{G}^{11,a}\p^{b}\mat{D}^c \nonumber \\
    &- \mat{S}^{abc}\mat{W}
    - \mat{S}^{ab}\mat{W}^{c}
    - \mat{S}^{ac}\mat{W}^{b}
    - \mat{S}^{a}\mat{W}^{bc} \nonumber
  \end{align}
\end{subequations}
and similarly for higher-order response functions.

The governing equations for the perturbed \acs{ASC} densities are
obtained in complete analogy with the handling of pertubed density
matrices outlined above.
We introduce a decomposition of the \acs{ASC} density into frequency
components into the perturbation-strength derivative of
Eq. \eqref{eq:polarization-stat}, so that
\begin{equation}\label{eq:PCM-differentiated}
  \V\p_\omega^{b_N} + \Tr \mathbfup{s}\mat{D}_\omega^{b_N} =
  \mathbb{S}^{(n-1)}_\omega
  \eqtr - \left(\mat{\varphi}\mat{D}\right)_{\omega, n-1}^{b_N} \text{.}
\end{equation}
$\mathbb{S}^{(n-1)}_\omega$ contains all terms that depend on lower order density matrices and
differentiated source integrals.
The term $\mathbb{S}^{(n-1)}_\omega$ always contains at least a first derivative of the
electrostatic potential integrals and is thus zero if the basis set is
independent of the perturbation tuple under consideration.
We now introduce a partition of the frequency components of the
polarization $\p_\omega^{b_N}$ into homogeneous (H) and particular (P)
terms:
\begin{equation}\label{eq:polarization-partition}
  \p_\omega^{b_N} = \p_\mathrm{P}^{b_N} + \p_\mathrm{H}^{b_N} \text{,}
\end{equation}
a similar partition exists for the density matrix and we thus get
the following system of equations:
\begin{subequations}
  \begin{align}
    \V\p_\mathrm{H}^{b_N} &+ \Tr \mathbfup{s}\Dh{b_N} = 0
    \label{eq:homo-polarization-eq}\\
    \V\p_\mathrm{P}^{b_N} &+ \Tr \mathbfup{s}\Dp{b_N}
    = \mathbb{S}^{(n-1)}_\omega \text{.}
 \label{eq:part-polarization-eq}
  \end{align}
\end{subequations}
We note that the particular term of the polarization is nonzero if and
only if the basis set depends on the external perturbation.

We finally turn our attention to the \acs{TDSCF} equation.
The perturbation-strength differentiated generalized \acs{KS} matrix is
first separated into its frequency components $\mathbfcal{F}^{b_N}_\omega$.
The H/P partition introduced for the variational densities will induce a
similar partition into these frequency components:
\begin{equation}\label{eq:generalized-Fock-partition}
  \mathbfcal{F}_\omega^{b_N} =
  \mat{G}^{\mathrm{KS}}(\Dh{b_N}) + \ASCh{b_N}\mat{\esp}
  + \breve{\mathbfcal{F}}_\omega^{b_N} \text{.}
\end{equation}
The two-electron and \acs{XC} contributions depending on the homogeneous
perturbed density matrix have been collected into the
$\mat{G}^\mathrm{KS}(\Dh{b_N})$ matrix, while all other contributions
are collected into $\breve{\mathbfcal{F}}_\omega^{b_N}$, and we
refer to \noparcite[refs.][]{Thorvaldsen2008-sg} and
\noparcite[][]{Ringholm2014-gx} for a more detailed treatment of this.
The parametrization of the homogeneous part of the perturbed density
matrix can be exploited to conveniently reformulate the perturbed
\acs{TDSCF} equation, so that
\begin{equation}\label{eq:response-equation}
 \left[\genHessian - \omega_{b_N}\genMetric \right]\rspParam{b_N} = \RHS{b_N}
\end{equation}
where the generalized Hessian $\genHessian$ and metric $\genMetric$
matrices were introduced and are defined by their transformations on the
response parameters $\rspParam{b_N}$:\autocite{Larsen2000-hj,
Kjaergaard2008-hy}
\begin{align}
  \genHessian\rspParam{b_N} &=
  \mat{G}^\mathrm{KS}([\rspParam{b_N}, \mat{D}]_{\mat{S}})\mat{D}\mat{S}
  -\mat{S}\mat{D}\mat{G}^\mathrm{KS}([\rspParam{b_N}, \mat{D}]_{\mat{S}})
+\mat{F} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S}
-\mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{F}
+ \ASCh{b_N}\mat{\esp}\mat{D}\mat{S}
-\mat{S}\mat{D}\mat{\esp}\ASCh{b_N}
\label{eq:E2-lintra}
\\
  \genMetric\rspParam{b_N} &= \mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S} \text{.}
  \label{eq:S2-lintra}
\end{align}
The generalized Hessian matrix $\genHessian$ includes two types of
solvent term. \emph{Implicit} terms are those included in the
zeroth-order Fock matrix, $\mat{F}$.
The last two terms in Eq. \ref{eq:E2-lintra} are \emph{explicit} terms,
as they explicitly involve the $N$-th order homogenous \acs{ASC}
variational density.
The theoretical treatment of frequency-dependent properties in solution
within the \acs{PCM} requires adoption of a nonequilibrium response
framework.\autocite{Cammi1999-rb, Tomasi2005-vm}
The explicit \acs{PCM} terms in Eq. \ref{eq:E2-lintra} are then
evaluated using the optical permittivity $\dield$ of the solvent
instead of the static permittivity $\diels$ used to compute the
implicit contributions.
In contrast to $\genHessian$, the generalized metric matrix --
$\genMetric$ -- is unchanged.
The \ac{RHS} in the response equation only includes terms that
depend on particular contributions up to the desired order or lower-order perturbed density
matrices:
\begin{equation}\label{eq:RHS}
  \RHS{b_N} =
  \left[\left( \tilde{mathbfcal{F}}
-
\frac{\mathrm{i}}{2}\pertmat{S}\deriv{}{t}\right)\pertmat{D}\pertmat{S}\right]^{\ominus,
b_N}_\mathrm{P}
\end{equation}

Solution strategy for response equations~\autocite{Saad2003-oa,
Saad2011-gm, Kauczor2011-rd, Malmqvist2013-vw}
